---
title: 'Impact of Weather Events on Population Health and Economy'
author: "Ryan Lei"
date: "September 21, 2014"
output:
  html_document:
    keep_md: true
---
Peer Assessment 2 of the course Reproducible Research in Data Science Specialization @ Coursera

## Synopsis
In this report, we explore the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database, whose data were collected from 1950 to 2011. We then investigate the impact of harmful weather events on population health and economy by summarizing the injuries, fatalities, property damage, and crop damage caused by each type of weather events. From this analysis, we conclude that tornadoes are the most harmful to population health, and floods cause the greatest economic damage. We present these findings in two stacked bar plots.

## Data Processing
### Load R Packages
```{r}
library(dplyr)
library(reshape2)
library(ggplot2)
library(grid)
```

### Reading Data
First, we read the raw CSV file as a data frame. The `read.csv()` function can accept .bz2 files without having to explicitly decompress it. This step is very time-consuming so we will **cache** this code chunk.
```{r, cache = TRUE}
raw_data <- read.csv(bzfile("repdata-data-StormData.csv.bz2"), stringsAsFactors = F)
```
As the initial investigation, we look at the dimension and column names.
```{r}
dim(raw_data)
names(raw_data)
```
### Subsetting
Since we are interested in the **overall** impact of weather events on population health and economy, regardless of the *year* and *state* in which the events occurred, we will only use the columns of `EVTYPE`, `FATALITIES`, `INJURIES`, `PROPDMG`, `PROPDMGEXP`, `CROPDMG`, and `CROPDMGEXP`.
```{r}
sub_data <- raw_data[, c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
```
Then, we subset only those rows whose `FATALITIES`, `INJURIES`, `PROPDMG`, and `CROPDMG` are not all zeros. This results in less than 30% of the original rows.
```{r}
sub_data <- subset(sub_data, FATALITIES + INJURIES + PROPDMG + CROPDMG > 0)
dim(sub_data)
```

### Data Transformations of Numbers
The power-of-10 exponents for property damange and crop damage (`PROPDMGEXP` and `CROPDMGEXP`) are messy, so we need to convert them into numbers (`NUM_PROPDMGEXP` and `NUM_CROPDMGEXP`).
```{r}
table(sub_data$PROPDMGEXP)
table(sub_data$CROPDMGEXP)
```
```{r}
# numbers (0, 2, 3, 4, 5, 6, 7) -> as is
sub_data <- mutate(sub_data,
                   NUM_PROPDMGEXP = as.numeric(PROPDMGEXP),
                   NUM_CROPDMGEXP = as.numeric(CROPDMGEXP))
# ("", -, +, ?) -> 0
sub_data$NUM_PROPDMGEXP[which(sub_data$PROPDMGEXP %in% c("", "+", "-", "?"))] <- 0
sub_data$NUM_CROPDMGEXP[which(sub_data$CROPDMGEXP %in% c("", "+", "-", "?"))] <- 0
# (H, h) -> hundred -> 2
sub_data$NUM_PROPDMGEXP[which(sub_data$PROPDMGEXP %in% c("H", "h"))] <- 2
sub_data$NUM_CROPDMGEXP[which(sub_data$CROPDMGEXP %in% c("H", "h"))] <- 2
# (K, k) -> kilo -> 3
sub_data$NUM_PROPDMGEXP[which(sub_data$PROPDMGEXP %in% c("K", "k"))] <- 3
sub_data$NUM_CROPDMGEXP[which(sub_data$CROPDMGEXP %in% c("K", "k"))] <- 3
# (M, m) -> million -> 6
sub_data$NUM_PROPDMGEXP[which(sub_data$PROPDMGEXP %in% c("M", "m"))] <- 6
sub_data$NUM_CROPDMGEXP[which(sub_data$CROPDMGEXP %in% c("M", "m"))] <- 6
# (B, b) -> billion -> 9
sub_data$NUM_PROPDMGEXP[which(sub_data$PROPDMGEXP %in% c("B", "b"))] <- 9
sub_data$NUM_CROPDMGEXP[which(sub_data$CROPDMGEXP %in% c("B", "b"))] <- 9
```
The numeric power-of-10 exponents after cleaning:
```{r}
table(sub_data$NUM_PROPDMGEXP, useNA = "always")
table(sub_data$NUM_CROPDMGEXP, useNA = "always")
```
Convert the property damage and crop damage into actual values:
```{r}
sub_data <- mutate(sub_data,
                   VAL_PROPDMG = PROPDMG * 10 ^ NUM_PROPDMGEXP,
                   VAL_CROPDMG = CROPDMG * 10 ^ NUM_CROPDMGEXP)

```
Finally, keep only the meaningful columns and drop intermediate calculations.
```{r}
final_data <- sub_data[, c("EVTYPE", "FATALITIES", "INJURIES", "VAL_PROPDMG", "VAL_CROPDMG")]
```

### Data Transformations of Event Types
There are way numerous unique values in `EVTYPE`.
```{r}
length(unique(final_data$EVTYPE))
```
As a *simple* cleaning, we will lowercase the letters and replace all the blank and punctuation characters with a space.
```{r}
final_data$EVTYPE <- tolower(final_data$EVTYPE)
final_data$EVTYPE <- gsub("[[:blank:][:punct:]]+", " ", final_data$EVTYPE)
# Trim the beginning and trailing spaces
final_data$EVTYPE <- gsub("^ ", "", final_data$EVTYPE)
final_data$EVTYPE <- gsub(" $", "", final_data$EVTYPE)
length(unique(final_data$EVTYPE))
```

## Results
### Question 1: Across the United States, which types of events are most harmful with respect to population health?
Summarize the injuries and fatalities by each type of weather events. Also, melt the data frame for the use of stacked bar plots in {ggplot2}.
```{r}
topN <- 10
result_health <- mutate(summarize(group_by(final_data, EVTYPE),
                                  injuries = sum(INJURIES),
                                  fatalities = sum(FATALITIES)),
                        fatalities_PLUS_injuries = injuries + fatalities)
result_health <- head(arrange(result_health, desc(fatalities_PLUS_injuries)), topN)
result_health_melt <- melt(result_health[, 1:3], id.vars = "EVTYPE", variable.name = "DamageType", value.name = "count")
```

Show and plot the top 10 weather events that cause the most fatalities or injuries.
```{r, fig.width = 9}
print(as.data.frame(result_health))
ggplot(result_health_melt, aes(x = reorder(EVTYPE, count), y = count, fill = DamageType)) + geom_bar(stat = "identity") + 
    coord_flip() + labs(title = "Impact of Weather Events on Population Health",
                        x = "Weather Event", y = "Number of Fatalities and Injuries", fill = "Damage Type") +
    theme(text = element_text(size = 14), plot.title = element_text(face = "bold"), 
          axis.title.x = element_text(vjust = -0.5), axis.title.y = element_text(vjust = +0.5), 
          title = element_text(vjust = +1.5))
```

**Answer 1: Significantly, tornadoes are the most harmful event to population health. Other harmful events include excessive heat, tstm wind, floods, and lightning.**

### Question 2: Across the United States, which types of events have the greatest economic consequences?
Summarize the property damage and crop damage by each type of weather events. Also, melt the data frame for the use of stacked bar plots in {ggplot2}.
```{r}
result_econ <- mutate(summarize(group_by(final_data, EVTYPE),
                                property = sum(VAL_PROPDMG),
                                crop = sum(VAL_CROPDMG)),
                      property_PLUS_crop = property + crop)
result_econ <- head(arrange(result_econ, desc(property_PLUS_crop)), topN)
result_econ_melt <- melt(result_econ[, 1:3], id.vars = "EVTYPE", variable.name = "DamageType", value.name = "count")
```

Show and plot the top 10 weather events that cause the most fatalities or injuries.
```{r, fig.width = 9}
print(format(as.data.frame(result_econ), digits = 3))
ggplot(result_econ_melt, aes(x = reorder(EVTYPE, count), y = count, fill = DamageType)) + geom_bar(stat = "identity") + 
    coord_flip() + labs(title = "Impact of Weather Events on Economy",
                        x = "Weather Event", y = "Damage on Property and Crop (US Dollars)", fill = "Damage Type") +
    theme(text = element_text(size = 14), plot.title = element_text(face = "bold"),
          axis.title.x = element_text(vjust = -0.5), axis.title.y = element_text(vjust = +0.5),
          title = element_text(vjust = +1.5))
```

**Answer 2: Significantly, floods cause the greatest economic damage. Other economically damaging events include hurricane typhoons, tornadoes, and storm surge.**

## References
1. NOAA Storm Database (CSV file in bzip2 compression): https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2
2. National Weather Service Storm Data Documentation: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
3. National Climatic Data Center Storm Events FAQ: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf
