---
title: "Statistical Inference -- Part 2"
author: "[Ryan Lei](https://www.coursera.org/user/i/adc3888c696d25d769137c26e7006fd8)"
date: "January 3, 2015"
output: html_document
---
### Introduction
In this exercise, we will perform a basic inferential data analysis on the **ToothGrowth** dataset built in R.

### Exploratory Data Analysis
From the help page of command `?ToothGrowth`, we know that this dataset measures the tooth lengths of 10 Guinea pigs grouped by **2 supplement types**: orange juice (OJ) and ascorbic acid (VC), and **3 dose levels**: 0.5, 1, 2 mg.

Let's see what the dataset looks like:
```{r}
head(ToothGrowth, 3)
# Count the number of observations in each group
with(ToothGrowth, table(supp, dose))
```

Indeed, there are 10 observations in each of the 6 groups: 2 supplement types * 3 dose levels.

Let's compare the *means* by summarizing each group:
```{r}
xtabs(len ~ supp + dose, aggregate(len ~ supp + dose, ToothGrowth, mean))
```

Presented graphically, here's a conditioning plot provided in the help page, with the **red** lines being the means in the above table:
```{r, fig.align='center'}
require(graphics)
coplot(len ~ dose | supp, data = ToothGrowth, panel = panel.smooth,
       xlab = "ToothGrowth data: length vs dose, given type of supplement")
```

### Intuition
At first glance, it seems OJ supplement is more effective than VC. In addition, dose level has a positive effect on tooth lengths. However, at a high dose level (from 1.0 to 2.0), this effect is not as strong. Even the situation of OJ outperforming VC disappears at dose = 2.0.

Next, we will conduct confidence intervals and hypothesis testing to verify this intuition.

### Confidence Interval and Hypothesis Testing
In this exercise, we will assume the 10 pigs are *NOT paired*, i.e., the pig #1 in one group is different from the pig #1 in another group. Therefore, they are treated as **independent groups**. Also, we will assume **unequal variances** across groups because we are not given such information.

First, we subset the dataset into different variables for easier references.
```{r}
OJ_05 <- subset(ToothGrowth, supp == "OJ" & dose == 0.5)$len
OJ_10 <- subset(ToothGrowth, supp == "OJ" & dose == 1.0)$len
OJ_20 <- subset(ToothGrowth, supp == "OJ" & dose == 2.0)$len
VC_05 <- subset(ToothGrowth, supp == "VC" & dose == 0.5)$len
VC_10 <- subset(ToothGrowth, supp == "VC" & dose == 1.0)$len
VC_20 <- subset(ToothGrowth, supp == "VC" & dose == 2.0)$len
```

In this exercise, we set all significance levels $\alpha$ to 5% to derive 95% confidence intervals. Because the exploratory data analysis showed a nice intuition that generally OJ > VC and higher dose > lower dose, all the comparisons will be done in **one-sided T tests** on differences in group means.

**We will reject the null hypothesis if *zero* is outside the 95% confidence interval, or equivalently, p-value < 0.05.**

**(1-1) At dose = 0.5 --------- $H_0$: $\mu_{OJ}$ = $\mu_{VC}$, $H_a$: $\mu_{OJ}$ > $\mu_{VC}$ **

```{r}
test11 <- t.test(OJ_05, VC_05, alternative = "greater", paired = F, var.equal = F)
list(confidence.interval = test11$conf.int, p.value = test11$p.value)
```
* Result: Reject $H_0$ in favor of $H_a$.

**(1-2) At dose = 1.0 --------- $H_0$: $\mu_{OJ}$ = $\mu_{VC}$, $H_a$: $\mu_{OJ}$ > $\mu_{VC}$ **

```{r}
test12 <- t.test(OJ_10, VC_10, alternative = "greater", paired = F, var.equal = F)
list(confidence.interval = test12$conf.int, p.value = test12$p.value)
```
* Result: Reject $H_0$ in favor of $H_a$.

**(1-3) At dose = 2.0 --------- $H_0$: $\mu_{OJ}$ = $\mu_{VC}$, $H_a$: $\mu_{OJ}$ > $\mu_{VC}$ **

```{r}
test13 <- t.test(OJ_20, VC_20, alternative = "greater", paired = F, var.equal = F)
list(confidence.interval = test13$conf.int, p.value = test13$p.value)
```
* Result: **Do not** reject $H_0$.

--------

**(2-1) At supp = OJ --------- $H_0$: $\mu_{0.5}$ = $\mu_{1.0}$, $H_a$: $\mu_{0.5}$ < $\mu_{1.0}$ **
```{r}
test21 <- t.test(OJ_05, OJ_10, alternative = "less", paired = F, var.equal = F)
list(confidence.interval = test21$conf.int, p.value = test21$p.value)
```
* Result: Reject $H_0$ in favor of $H_a$.

**(2-2) At supp = VC --------- $H_0$: $\mu_{0.5}$ = $\mu_{1.0}$, $H_a$: $\mu_{0.5}$ < $\mu_{1.0}$ **
```{r}
test22 <- t.test(VC_05, VC_10, alternative = "less", paired = F, var.equal = F)
list(confidence.interval = test22$conf.int, p.value = test22$p.value)
```
* Result: Reject $H_0$ in favor of $H_a$.

**(2-3) At supp = OJ --------- $H_0$: $\mu_{1.0}$ = $\mu_{2.0}$, $H_a$: $\mu_{1.0}$ < $\mu_{2.0}$ **
```{r}
test23 <- t.test(OJ_10, OJ_20, alternative = "less", paired = F, var.equal = F)
list(confidence.interval = test23$conf.int, p.value = test23$p.value)
```
* Result: **Do not** reject $H_0$.

**(2-4) At supp = VC --------- $H_0$: $\mu_{1.0}$ = $\mu_{2.0}$, $H_a$: $\mu_{1.0}$ < $\mu_{2.0}$ **
```{r}
test24 <- t.test(VC_10, VC_20, alternative = "less", paired = F, var.equal = F)
list(confidence.interval = test24$conf.int, p.value = test24$p.value)
```
* Result: Reject $H_0$ in favor of $H_a$.

### Conclusions
At first glance, it seems OJ supplement is more effective than VC. In addition, dose level has a positive effect with the tooth length. However, at high dose level of 2.0, the effect is not as strong. Even the situation of OJ outperforming VC disappear at dose = 2.0.

Assumptions: independent groups
